---
title: "Getting Started with HybPhaser"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with HybPhaser}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(HybPhaser)
```

## Introduction

HybPhaser is designed to detect and phase hybrid accessions in target 
capture datasets from polyploid species. It works as an extension to the 
[HybPiper](https://github.com/mossmatters/HybPiper) assembly pipeline.

The core workflow involves:

1. **SNP Assessment** - Detect hybrids by measuring heterozygosity in 
   consensus sequences
2. **Clade Association** - Map reads to divergent reference clades
3. **Phasing** - Separate reads according to similarity with parental 
   clades
4. **Sequence List Generation** - Create merged datasets for 
   phylogenetic analysis

This vignette demonstrates the package workflow using the functions 
currently available. Additional functionality is being added as the 
package conversion continues.

## Prerequisites

HybPhaser expects that you have already run HybPiper on your target 
capture data. You need:

- HybPiper output directory containing assembled sequences for each sample
- A FASTA file with target sequences (baits)
- A namelist file with sample names (one per line)

### External Dependencies

Several command-line tools are required for the complete workflow:

- **BWA** (v0.0.17+) - for read mapping
- **SAMtools** (v1.9+) - for processing alignments
- **Bcftools** (v1.9+) - for variant calling
- **BBMap/BBSplit** (v38.87+) - for clade association and phasing

## Step 1: Generate Consensus Sequences

Before using the R functions, you need to generate consensus sequences 
from your HybPiper output. HybPhaser includes bash scripts that handle 
this step.

### Accessing the Scripts

```{r}
# Get path to the consensus sequence generation script
script <- hybphaser_scripts("1_generate_consensus_sequences.sh")
script

# List all available scripts
list.files(hybphaser_scripts())
```

### Running the Script

The script can be run from the command line or from within R:

```{r eval=FALSE}
# Preview the command (doesn't actually run it)
run_hybphaser_script(
  "1_generate_consensus_sequences.sh",
  args = c(
    "-n", "path/to/namelist.txt",
    "-p", "path/to/hybpiper_output",
    "-o", "path/to/hybphaser_output",
    "-d", "10",    # Minimum coverage depth
    "-f", "0.15",  # Minimum allele frequency
    "-a", "4"      # Minimum allele count
  ),
  dry_run = TRUE
)
```

This script:
- Collects reads and contigs from HybPiper output
- Maps reads to contigs using BWA
- Calls variants with Bcftools
- Generates consensus sequences with ambiguity codes for heterozygous 
  sites

## Step 2: Count SNPs in Consensus Sequences

Once consensus sequences are generated, we can count SNPs (polymorphic 
sites marked as ambiguity codes) across samples and loci.

### Setting Up

First, let's create a minimal example to demonstrate the workflow:

```{r}
# Create temporary directory structure
temp_dir <- tempfile()
dir.create(file.path(temp_dir, "01_data", "sample1", "consensus"), 
           recursive = TRUE)

# Create a simple target file
target_file <- file.path(temp_dir, "targets.fasta")
writeLines(c(
  ">species1-gene1",
  "ATCGATCGATCGATCGATCG",
  ">species1-gene2", 
  "GCTAGCTAGCTAGCTAGCTA"
), target_file)

# Create consensus sequences with some SNPs
# Gene1: has SNPs (R and Y are ambiguity codes)
writeLines(c(
  ">gene1",
  "ATCGATRYATCGATCGATCG"
), file.path(temp_dir, "01_data", "sample1", "consensus", "gene1.fasta"))

# Gene2: no SNPs (clean sequence)
writeLines(c(
  ">gene2",
  "GCTAGCTAGCTAGCTAGCTA"
), file.path(temp_dir, "01_data", "sample1", "consensus", "gene2.fasta"))

# Create namelist
namelist <- file.path(temp_dir, "namelist.txt")
writeLines("sample1", namelist)
```

### Running SNP Count

```{r}
# Count SNPs across all samples and loci
results <- count_snps(
  path_to_output_folder = temp_dir,
  fasta_file_with_targets = target_file,
  targets_file_format = "DNA",
  path_to_namelist = namelist,
  intronerated_contig = FALSE
)
```

### Understanding the Results

The function returns two data frames:

```{r}
# SNP proportions per locus and sample
results$tab_snps

# Sequence lengths per locus and sample  
results$tab_length
```

The SNP proportion indicates the level of heterozygosity:

- **Low values (< 0.01)** suggest non-hybrid samples or homozygous 
  regions
- **High values (> 0.02)** may indicate hybrids or paralogous genes
- **Very high values (> 0.05)** often suggest assembly issues or 
  contamination

### Saved Output

Results are automatically saved as R objects for downstream analysis:

```{r}
# Check what was saved
list.files(file.path(temp_dir, "00_R_objects"), recursive = TRUE)
```

These can be loaded in subsequent steps:

```{r eval=FALSE}
# Load saved results
snps <- readRDS(file.path(temp_dir, "00_R_objects", "Table_SNPs.Rds"))
lengths <- readRDS(file.path(temp_dir, "00_R_objects", 
                             "Table_consensus_length.Rds"))
```

## Working with Real Data

With real HybPiper data, the workflow would look like:

```{r eval=FALSE}
# Count SNPs in your actual data
my_results <- count_snps(
  path_to_output_folder = "hybphaser_output",
  fasta_file_with_targets = "my_targets.fasta",
  targets_file_format = "DNA",
  path_to_namelist = "my_samples.txt",
  intronerated_contig = FALSE  # Set TRUE if using intronerate
)

# Examine results
head(my_results$tab_snps)
summary(my_results$tab_snps)

# Identify samples with high heterozygosity (potential hybrids)
sample_heterozygosity <- colMeans(my_results$tab_snps, na.rm = TRUE)
potential_hybrids <- names(sample_heterozygosity[sample_heterozygosity > 0.02])
print(potential_hybrids)

# Identify loci with high SNP rates (potential paralogs)
locus_heterozygosity <- rowMeans(my_results$tab_snps, na.rm = TRUE)
potential_paralogs <- names(locus_heterozygosity[locus_heterozygosity > 0.02])
print(potential_paralogs)
```

## Using Example Data

HybPhaser includes example configuration files:

```{r}
# Access example clade references
clade_refs <- read.csv(hybphaser_example("clade_references.csv"))
clade_refs

# Access example phasing preparation file
phasing_prep <- read.csv(hybphaser_example("phasing_prep.csv"))
phasing_prep
```

## Next Steps

After counting SNPs, the typical HybPhaser workflow continues with:

1. **Dataset Assessment** - Filter samples and loci based on 
   heterozygosity
2. **Sequence List Generation** - Create FASTA files for alignment
3. **Clade Association** - Map reads to reference clades using BBSplit
4. **Phasing** - Separate hybrid reads into parental haplotypes
5. **Merging** - Combine phased and non-phased sequences

These additional functions are being converted to the package format and 
will be documented in future vignette updates.

## Getting Help

- Function documentation: `?count_snps`
- Report issues: <https://github.com/LarsNauheimer/HybPhaser/issues>
- Original HybPhaser paper: [bioRxiv preprint](
  https://www.biorxiv.org/content/10.1101/2020.10.27.354589v2)

## Clean Up

```{r}
# Remove temporary files
unlink(temp_dir, recursive = TRUE)
```
