---
title: "HybPhaser Example Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{HybPhaser Example Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(HybPhaser)

# Check if Docker is available for this vignette
has_docker <- check_docker(quiet = TRUE)
has_image <- FALSE
if (has_docker) {
  has_image <- check_docker_image("joelnitta/hybphaser:latest", 
                                   pull = FALSE)
}
```

## Introduction

HybPhaser is designed to detect and phase hybrid accessions in target 
capture datasets from polyploid species. It works as an extension to the 
[HybPiper](https://github.com/mossmatters/HybPiper) assembly pipeline.

The core workflow involves:

1. **SNP Assessment** - Detect hybrids by measuring heterozygosity in 
   consensus sequences
2. **Clade Association** - Map reads to divergent reference clades
3. **Phasing** - Separate reads according to similarity with parental 
   clades
4. **Sequence List Generation** - Create merged datasets for 
   phylogenetic analysis

This vignette demonstrates the package workflow using the functions 
currently available. Additional functionality is being added as the 
package conversion continues.

## Example with Test Data

HybPhaser can download the official HybPiper test dataset on demand. You can
run HybPiper on that data via conda, then use the resulting real HybPiper
output with HybPhaser.

The example below is the full end-to-end workflow:
1. Run HybPiper (conda)
2. Generate consensus sequences (Docker)
3. Count SNPs (R)

If HybPiper is installed in a conda environment (default name:
`hybpiper_env`), you can download and run the official test dataset and
continue directly with HybPhaser:

```{r eval=FALSE, echo=TRUE}
# 0. Optional checks
check_conda()
check_hybpiper_conda("hybpiper_env")
check_docker()
check_docker_image("joelnitta/hybphaser:latest", pull = TRUE)

# 1. Download and run official HybPiper test data using conda
hp_data <- run_hybpiper_test_dataset(
  output_dir = tempfile("hybpiper_test_"),
  samples = c("EG30"),
  conda_env = "hybpiper_env",
  cpu = 1,
  mapper = "bwa"
)

# 2. Generate consensus sequences with HybPhaser Docker workflow
run_generate_consensus_sequences(
  hybpiper_dir = hp_data$hybpiper_dir,
  output_dir = file.path(hp_data$hybpiper_dir, "hybphaser_output"),
  namelist = hp_data$namelist,
  threads = 1
)

# 3. Count SNPs
snp_results <- count_snps(
  path_to_output_folder = file.path(hp_data$hybpiper_dir, "hybphaser_output"),
  fasta_file_with_targets = hp_data$targets_file,
  targets_file_format = "DNA",
  path_to_namelist = hp_data$namelist
)

# 4. Inspect results
head(snp_results$tab_snps)
head(snp_results$tab_length)
```

## Next Steps

After counting SNPs, the typical HybPhaser workflow continues with:

1. **Dataset Assessment** - Filter samples and loci based on 
   heterozygosity
2. **Sequence List Generation** - Create FASTA files for alignment
3. **Clade Association** - Map reads to reference clades using BBSplit
4. **Phasing** - Separate hybrid reads into parental haplotypes
5. **Merging** - Combine phased and non-phased sequences

These additional functions are being converted to the package format and 
will be documented in future vignette updates.

## Getting Help

- Function documentation: `?count_snps`
- Report issues: <https://github.com/LarsNauheimer/HybPhaser/issues>
- Original HybPhaser paper: [bioRxiv preprint](
  https://www.biorxiv.org/content/10.1101/2020.10.27.354589v2)

## Clean Up

```{r}
# Remove temporary files
unlink(temp_dir, recursive = TRUE)
```
