---
title: "HybPhaser Example Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{HybPhaser Example Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(HybPhaser)

# Check if Docker is available for this vignette
has_docker <- check_docker(quiet = TRUE)
has_image <- FALSE
if (has_docker) {
  has_image <- check_docker_image("joelnitta/hybphaser:latest",
    pull = FALSE
  )
}
```

## Introduction

HybPhaser is designed to detect and phase hybrid accessions in target 
capture datasets from polyploid species. It works as an extension to the 
[HybPiper](https://github.com/mossmatters/HybPiper) assembly pipeline.

The core workflow involves:

1. **SNP Assessment** - Detect hybrids by measuring heterozygosity in 
   consensus sequences
2. **Dataset Quality Assessment** - Filter samples and loci based on 
   missing data and heterozygosity
3. **Sequence List Generation** - Create organized FASTA files for loci 
   and samples
4. **Clade Association** - Map reads to divergent reference clades
5. **Phasing** - Separate reads according to similarity with parental 
   clades

This vignette demonstrates the package workflow using the functions 
currently available. Additional functionality is being added as the 
package conversion continues.

## Example with Test Data

HybPhaser can download the official HybPiper test dataset on demand. You can
run HybPiper on that data via conda, then use the resulting real HybPiper
output with HybPhaser.

The example below is the full end-to-end workflow:
1. Run HybPiper (conda)
2. Generate consensus sequences (Docker)
3. Count SNPs (R)
4. Assess dataset quality (R)
5. Generate sequence lists (R)
6. Extract mapped reads for clade association (R)
7. Prepare/run BBSplit script for clade association (R)
8. Collate BBSplit results (R)

If HybPiper is installed in a conda environment (default name:
`hybpiper_env`), you can download and run the official test dataset and
continue directly with HybPhaser:

```{r eval=FALSE, echo=TRUE}
# 0. Optional checks
check_conda()
check_hybpiper_conda("hybpiper_env")
check_docker()
check_docker_image("joelnitta/hybphaser:latest", pull = TRUE)

# 1. Download and run official HybPiper test data using conda
hp_data <- run_hybpiper_test_dataset(
  output_dir = tempfile("hybpiper_test_"),
  samples = c("EG30"),
  conda_env = "hybpiper_env",
  cpu = 1,
  mapper = "bwa"
)

# 2. Generate consensus sequences with HybPhaser Docker workflow
run_generate_consensus_sequences(
  hybpiper_dir = hp_data$hybpiper_dir,
  output_dir = file.path(hp_data$hybpiper_dir, "hybphaser_output"),
  namelist = hp_data$namelist,
  threads = 1
)

# 3. Count SNPs
snp_results <- count_snps(
  path_to_output_folder = file.path(
    hp_data$hybpiper_dir,
    "hybphaser_output"
  ),
  fasta_file_with_targets = hp_data$targets_file,
  targets_file_format = "DNA",
  path_to_namelist = hp_data$namelist
)

# 4. Assess dataset quality
# Generate tables and plots to assess sequence recovery and identify
# putative paralogs
assessment <- assess_dataset(
  snp_table = snp_results$tab_snps,
  length_table = snp_results$tab_length,
  targets_file = hp_data$targets_file,
  targets_type = "DNA",
  output_dir = file.path(hp_data$hybpiper_dir, "hybphaser_output"),
  min_loci_per_sample_prop = 0.5,
  min_samples_per_locus_prop = 0.5,
  paralog_threshold = "outliers",
  remove_sample_outliers = TRUE
)

# 5. Generate sequence lists
# Create FASTA files organized by locus and by sample, applying quality
# filters from dataset assessment
seq_lists <- generate_sequence_lists(
  path_to_output_folder = file.path(
    hp_data$hybpiper_dir,
    "hybphaser_output"
  ),
  fasta_file_with_targets = hp_data$targets_file,
  targets_file_format = "DNA",
  path_to_namelist = hp_data$namelist
)

# 6. Extract mapped reads for clade association
mapped_reads <- extract_mapped_reads(
  base_dir = file.path(hp_data$hybpiper_dir, "hybphaser_output"),
  output_dir = file.path(
    hp_data$hybpiper_dir,
    "hybphaser_output",
    "mapped_reads"
  ),
  namelist = hp_data$namelist,
  remove_duplicate_sequences = FALSE
)

# 7. Inspect results
head(snp_results$tab_snps)
head(snp_results$tab_length)
head(assessment$summary_table)
cat("\nSequence lists created:\n")
cat("  Loci:", seq_lists$n_loci_files, "\n")
cat("  Samples:", seq_lists$n_sample_files, "\n")
cat("\nMapped reads files created:", mapped_reads$n_samples, "\n")

# 8. Prepare BBSplit clade-association script
# Requires a clade reference CSV with two columns: sample and abbreviation.
# Here we derive a minimal example directly from generated sample sequences.
reference_dir <- file.path(
  hp_data$hybpiper_dir,
  "hybphaser_output",
  "03_sequence_lists",
  "samples_consensus"
)

reference_files <- list.files(reference_dir, pattern = "_consensus\\.fasta$")
reference_samples <- sub("_consensus\\.fasta$", "", reference_files)

clade_reference_csv <- file.path(
  hp_data$hybpiper_dir,
  "hybphaser_output",
  "04_clade_association",
  "clade_references_example.csv"
)
dir.create(dirname(clade_reference_csv), recursive = TRUE, showWarnings = FALSE)

utils::write.csv(
  data.frame(
    samples = reference_samples,
    abb = paste0("C", seq_along(reference_samples))
  ),
  clade_reference_csv,
  row.names = FALSE
)

bbsplit_job <- prepare_bbsplit_script(
  path_to_clade_association_folder = file.path(
    hp_data$hybpiper_dir,
    "hybphaser_output",
    "04_clade_association"
  ),
  csv_file_with_clade_reference_names = clade_reference_csv,
  path_to_reference_sequences = reference_dir,
  path_to_read_files_cladeassociation = file.path(
    hp_data$hybpiper_dir,
    "hybphaser_output",
    "mapped_reads"
  ),
  read_type_cladeassociation = "single-end",
  run_clade_association_mapping_in_R = TRUE,
  docker_fallback = TRUE
)

cat("\nBBSplit script prepared:", bbsplit_job$script_path, "\n")

# 9. Collate BBSplit results
clade_assoc <- collate_bbsplit_results(
  path_to_clade_association_folder = file.path(
    hp_data$hybpiper_dir,
    "hybphaser_output",
    "04_clade_association"
  ),
  csv_file_with_clade_reference_names = clade_reference_csv,
  path_to_output_folder = file.path(
    hp_data$hybpiper_dir,
    "hybphaser_output"
  )
)

head(clade_assoc$table_clade_association)
```

## Next Steps

After generating sequence lists and mapped read files, the typical HybPhaser
workflow continues with:

1. **Clade Association** - Map reads to reference clades using BBSplit
2. **Phasing** - Separate hybrid reads into parental haplotypes
3. **Merging** - Combine phased and non-phased sequences

These additional functions are being converted to the package format and 
will be documented in future vignette updates.

## Getting Help

- Function documentation: `?count_snps`, `?assess_dataset`,
  `?generate_sequence_lists`, `?extract_mapped_reads`,
  `?prepare_bbsplit_script`, `?collate_bbsplit_results`
- Report issues: <https://github.com/LarsNauheimer/HybPhaser/issues>
- Original HybPhaser paper: [bioRxiv preprint](
  https://www.biorxiv.org/content/10.1101/2020.10.27.354589v2)

